Manuscript_draft

### 2021-03-05 manuscript_draft.Rmd
k;j

To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects. To account for plot-level productivity differences, we included plot as a random intercept effect.  Models were fit using the 'lme4' package for R [@].  We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].  Pairwise contrasts were performed using the 'emmeans' package for R [@] and Tukey HSD method of p value adjustment was used to control for multiplicity.  We report all significant results at p<0.05.

@@

Stand density 
The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the two harvested treatments and control.  Increasing stand density reflects the growth of smaller trees that over time, achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.  The faster increases in stand density of the thinned treatments compared to the control likely reflect that smaller trees in the thinned stands responded with faster diameter increment after treatment (as previously discussed).  We note that our study did not account for new seedlings that may have become established over the period of the study.  Based on increasing densities in the smaller diameter classes, these stands will likely maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.  Thus, partial harvest treatments may be used in this context to maintain a structure typical of uneven-aged stands.

The low-removal thinning treatment also maintained a species composition similar to the unthinned stands, in that subalpine fir was the leading species in the smaller diameter classes, and spruce was the leading species in the larger classes.  This is a stand structure and composition common at low- to mid-elevations in the interior of British Columbia.  In closed forests with undisturbed forest floors, subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.  Compared to spruce, subalpine fir can be relatively short-lived, and can die after 250 years in the canopy.  Spruce can live for over 400 years and thus can achieve larger sizes and dominant canopy positions.  The high-removal unit had a somewhat different species composition.  Here, fir was the leading species in all diameter classes.  This means that the high-removal thinning treatment converted the stand composition to fir-leading, which may be a primary concern for many forest managers, as spruce is often preferred as a crop species over subalpine fir [].  However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and over the 27-year period of this study, spruce monotonically increased its composition (by volume and basal area) compared to fir.  Over time, we expect that spruce would continue to increase its proportion relative to fir and eventually will achieve a similar species composition - diameter distribution as the other units.

### Stand-level growth responses
Over the 27-year period of this study, plots with high-removal harvest treatments accrued basal area significantly faster than the control and low-removal harvest treatment, likely due to increases in tree-level growth and stand density discussed previously.  As a result, differences in basal area between the high-removal harvest unit and the control reduced from ca. 17m2/hectare in 1992 to ca.9m2/hectare in 2019.  In contrast, basal area increment between the low-removal unit and the control did not differ significantly over the 1992-2019 period.  The difference between these two units remained relatively constant over all measurement periods at ca. 10m2/hectare.  The similar rates of basal area increment between the low-removal thinning treatment and the control are difficult to explain.  Tree-level results showed that fir grew faster in the low-removal unit than the control, but spruce did not. As well, by 2019 there were no significant differences in stand density between the two units.  Thus, it appears that while the fir trees may have increased growth as a response of low-removal, the growth increases were insufficient to cause a significant change in basal area increment.  Furhter monitoring in this study will help determine the future trajectory of the low-removal thinning treatment.

While basal area increment differed between the high-removal treatment unit and the control, there was little difference in volume increment between any of the units.  A possible explanation is that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth; this would be important because the volume formula used in this study is based on both tree diameter and height [@Nigh2016].  It can be difficult to precisely measure tree height growth using field-based methods, particularly in mature trees, which may be a primary reason that studies tend to focus on radial growth responses to silviculture treatments and gap dynamics [@Vepakommaetal2011].  Commercial thinning treatments in spruce fir stands increased tree taper [@]    Enhanced radial growth is a common repsonse to reduced stand densities [@Plauborg2014].

Volume and basal area increment peaked in the 1994-1997 measurement period, with the strongest decline in annual growth over the last measurement period of 2009-2019.  Age-related declines in stand productivity may be associated with decreasing water-use efficiency and reduced photosynthesis, with the age of maximal growth rate varying with factors such as species and site conditions [@Ryan1997].  In both thinned treatments, maximal growth rates peaked in the 1997-2009 period.  Thus, it appears the thinning treatment may have delayed the period of maximal growth rates by approximately 10 years.  The timing of thinning treatments can be important.  Stands that are past their period of age-related maximal growth rate may not have as much capacity to respond to treatments as a younger stand, although this is liekly a more important factor for shade-intolerant species rather than the spruce and fir trees in this study [@Bose2017].

Stand mortality
Stand mortality increased with basal area after treatment.  The thinned stands had the lowest mortality over the time period studies.  Thinning is considered a silviculture treatment that improves growth efficiency of the stand by redistibuting growing space and capturing mortality that would otherwise reduce net stand growth.  Our findings suggest that this occured, and that the thinning treament captured a portion of the stand that would have been lost eventually to natural mortality.  This suggests that the thinning treatments may have captured some volume through trees that would have died naturally.  Thus, while volume removals from thinning are never entirely recoverable, this may not be true when mortality is considered.

[@Valkonen2020] found similar mortality in uneven-aged norway spruce stands.

>>>>

To test the hypothesis that tree-level radial growth response over the 1992-2019 period varied with interactions between (i) tree diameter at the time of treatment (1992); (ii) treatment; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 diameter increment as the response variable, and treatment, species and log-transformed 1992 diameter as interacting fixed effects. To account for plot-level productivity differences, we included plot as a random intercept effect.  Models were fit using the 'lme4' package for R [@].  We also fit a similar model with the 1992-2019 diameter increment expressed as a percentage of 1992 diameter.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@barton2009mumin].  Pairwise contrasts were performed using the 'emmeans' package for R [@] and Tukey HSD method of *p* value adjustment was used to control for multiplicity.  We report all significant results at p<0.05.

@@

Stand density 
The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the two harvested treatments and control.  This likely resulted from two processes; growth and mortality.  Increasing stand density reflects the growth of smaller trees that achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.  The faster increases in stand density of the thinned treatments compared to the control likely reflect that smaller trees in the thinned stands responded with faster diameter increment after treatment (as previously discussed).  We note that our study did not account for new seedlings that may have become established over the period of the study.  Based on increasing densities in the smaller diameter classes, these stands will likely maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.

  Stand densities between all units likely also converged because of different mortality rates.  Tree mortality increased with stand-level basal area, and the thinned stands had the lowest mortality over the time of the study.  Most of the mortality was concentrated in fir trees among the smaller size classes.  Density-dependent tree mortality is common in younger, even-aged stands, reflecting intertree competition for light and other resources, and resulting in increasing uniformity of tree distribution [@Larson2015].  In the uneven-aged stands of our study, density-dependent mortality likely reflects indicate the effects of assymetric competition and suppression of .  In the unthinned stands, smaller trees died more frequently as a result of suppression from the larger trees, whereas thinning reduced this competition.  Density-dependent mortality in uneven-aged, mixed-species forests can also vary with drought and site-level conditions such as aspect and soil moisture regime, which needs to be considered when planning stand thinning treatments [@VanGunst2016].

Species composition

The low-removal thinning treatment also maintained a species composition by diameter class similar to the unthinned stands.  In these units, subalpine fir was the leading species by frequency in the smaller diameter classes, and spruce was the leading species in the larger classes.  This is a stand structure and composition common at low- to mid-elevations in the interior of British Columbia.  In closed forests with undisturbed forest floors, subalpine fir can have higher seedling establishment success than spruce, leading to higher densities of fir in the understorey.  Compared to spruce, subalpine fir can be relatively short-lived, and can die after 250 years in the canopy.  Spruce can live for over 400 years and thus can achieve larger sizes and dominant canopy positions.

The high-removal unit had a somewhat different species composition by diameter class.  Here, fir was the leading species in all diameter classes.  This means that the high-removal thinning treatment converted the stand composition to fir-leading in all tree sizes.  This result may be concern for forest managers, as spruce is often preferred as a crop species over subalpine fir [].  However, the stands in this treatment had more spruce in the smaller diameter classes than the other treatment units and over the 27-year period of this study, spruce monotonically increased its composition (by volume and basal area) compared to fir.  Over time, we expect that spruce will continue to increase its proportion relative to fir and eventually achieve a similar species composition - diameter distribution as the other units.

### Tree mortality

, reflecting intertree competiton for light and other resources.  In older stands with more complex structures, tree mortality tends to be primarily caused by senescence and spatially autocorrelated forest agents.  Given that the forests in this study are older and have a stand structure typial of an uneven-aged stand,.  , which tends to result in spatially aggregated tree distributions but by Thinning is considered a silviculture treatment that improves growth efficiency of the stand by redistibuting growing space and capturing mortality that would otherwise reduce net stand growth.  Our findings support suggest that this occured, and that the thinning treament captured a portion of the stand that would have been lost eventually to natural mortality.  The thinning treatments may have captured some volume through trees that would have died naturally, potentially increasing the net recoverable volume from these stands.  [@Valkonen2020] found similar mortality in uneven-aged norway spruce stands.

### Stand-level growth responses
Over the 27-year period of this study, plots with high-removal harvest treatments accrued basal area significantly faster than the control and low-removal harvest treatment, likely due to increases in tree-level growth and stand density discussed previously.  As a result, differences in basal area between the high-removal harvest unit and the control reduced from ca. 17m2/hectare in 1992 to ca.9m2/hectare in 2019.

In contrast, basal area increment between the low-removal unit and the control did not differ significantly over the 1992-2019 period, and the difference between these two units remained relatively constant over all measurement periods at ca. 10m2/hectare.  The similar rates of basal area increment between the low-removal thinning treatment and the control are difficult to explain.  Tree-level results showed that fir grew faster in the low-removal unit than the control, but spruce did not. As well, by 2019 there were no significant differences in stand density between the two units.  Thus, it appears that while the fir trees may have increased growth as a response of low-removal, the growth increases were insufficient to cause a significant change in basal area increment.  Further monitoring in this study will help determine the future trajectory of the low-removal thinning treatment.

While our study found that high-removal partial harvest resulted in faster basal area growth, we did not find evidence that thinning changed volume increment in trees greater than 17.5cm DBH. A possible explanation is that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth; this would be important because the volume formula used in this study is based on both tree diameter and height [@Nigh2016].  We note that it is often difficult to precisely capture height growth changes in mature trees using conventional field-based methods [@Vepakommaetal2011], which makes capturing height changes problematic in these sorts of studies.  We also utilized a 17.5cm DBH limit for volume calculations; as such, our volume measures only reflect the contribution of the larger trees in the stand, and do not include volume increment in smaller trees.

The partial harvest treatment appeared to change the timing of peak stand growth.  In the control unit, Volume and basal area increment peaked in the 1994-1997 measurement period, and declined thereafter, with the strongest decline in annual growth over the last measurement period of 2009-2019.  Age-related declines in stand productivity may be associated with decreasing water-use efficiency and reduced photosynthesis, with the age of maximal growth rate varying with factors such as species and site conditions [@Ryan1997].  In both thinned treatments, maximal growth rates peaked in the 1997-2009 period.  Thus, it appears the thinning treatment may have delayed the period of maximal growth rates by approximately 10 years.  The timing of thinning treatments can be important.  Stands that are past their period of age-related maximal growth rate may not have as much capacity to respond to treatments as a younger stand, although this is liekly a more important factor for shade-intolerant species rather than the spruce and fir trees in this study [@Bose2017].  At the time of this study, the high-removal thinning treatment had maintianed an annual basal area and volume increment that was similar to previous measurement periods; if this high level of increment continues, differences in basal area and volume will continue to diminish compared to the control, and this stand may exceed the low-removal thinning treatment.

## Management of uneven-aged stands
